Semantic integration of physiology phenotypes with an application to the Cellular Phenotype Ontology

### 2012-04-04 paper/response-bioinformatics.txt

Dear Dr. Bateman,

Thank you very much for providing us with the opportunity to revise our
manuscript. We have now prepared a revision in which we have addressed
all reviewers' comments. In response to both reviewers, we have
removed the discussion of the heart beating example from the
introduction and now provide a discussion of the axioms in the
manuscript as well as a detailed presentation of the axioms as
supplementary material. We have included a response to the individual
points raised by the reviewers below.  We would like to thank the
reviewers for their valuable comments which allowed us to improve our
manuscript.

@@

Reviewer 1:
"Introduction: I find the heart beating example difficult to follow,
primarily because it lacks simple easy-to-understand definitions - so
either this section is removed or needs to be rewritten (the result
section, however, is much clearer)."

Response: We have removed the section on heart beating from the
introduction, as recommended by the reviewer.

Reviewer 1:
p4 - On the one hand, I agree with "abnormalities in cell division
resulting in Aneuploidy refer to abnormalities of cell division
processes, while an increased rate of cell division refers to an
abnormality in the pattern of occurrence of multiple cell division
processes". However, you then state "an increased rate of cell
division is not an attribute of cell division processes, but rather an
attribute of the regulation of cell division"
two questions: 
1. is the regulation of a process(es) not different from that
collection of processes?
2. is it not the case that a collection of processes has attributes
based on its parts?

Response:
We agree with the reviewer that "regulation" processes do not always
(or commonly) include the processes that are being regulated as
parts. We have modified our manuscript in this regard and we now provide
more details on regulation processes in GO and their relation to
collection processes.

@@

Reviewer 1:
This paper needs to discuss the axiom patterns that generates the ontology.

Response:
We agree with the reviewer and in response we now provide information about the
axioms both in the manuscript (new section 2.3 "Axioms for physiology
phenotypes"). We also provide the full details of all axioms as supplementary material.

@@

Reviewer 1:
abstract: suggests that its application is in the unification of
cellular phenotypes across species; but it's not clear where
species-specific differentiation would come into play. it also
suggests (and is further outlined in the discussion) that it could be
ussed for systems microscopy, but no evidence is provided to support
the claim. And the latter, really must be further developed, to
demonstrate that the ontology can be used for some application.

Response: 
The Cell Phenotype Ontology, which we have built by applying our
ontology-patterns for representing physiology phenotypes, is currently
being applied for the annotation of cellular observations in systems
microscopy. For this purpose, CPO provides the necessary terms (in the
sense of a controlled vocabulary) as well as a taxonomic structure
that is used for basic retrieval operations.

We agree with the reviewer that an application of our work in a
bioinformatics analysis would be highly beneficial. For example, once
sufficient data is annotated with the CPO, we can use its taxonomic
structure for enrichment analyses. Similarly, once more cellular
phenotypes are being annotated in various model organism databases, we
can utilize CPO for integrative phenotype analyses across
species. However, at the moment, no comprehensive ontology of
physiology phenotypes is available, partly due to the complexity of
the domain and partly due to limitations in available ontology
frameworks. We see our main contribution in the development of a
theory that allows to capture physiology phenotypes; we applied this
theory to construct (largely automatically) the CPO, which is now
being used to annotate cellular phenotypes in the systems microscopy
domain. Based on our contributions, it is now possible to annotate
and eventually integrate (through the use of a controlled vocabulary
of physiology phenotypes, of which the CPO provides a first example)
data in the domain of physiology. At the same time, we maintain
compatibility with GO (on which a majority of physiology phenotypes
are based) and other PATO-based phenotype ontologies, thereby filling
a gap that has limited data integration in the cellular physiology domain so
far.

@@

Reviewer 2:
A theory of physiology related phenotypes and an implementation based
on GO, GO cross-products and PATO is presented. An outline of a theory
of phenotypes is introduced which participate-in a set of pathological
processes which can be qualified by a small set of PATO qualities. At
least, for a majority of heterogeneous phenotype descriptions this can
be an approach for unification. In future work a sound formal
characterization has and an evaluation of the practical value of the
approach has to be delivered.

Response:
We have now added a detailed formal presentation of our axioms as
supplementary material. An evaluation of the practical utility depends
on data that is being annotated with physiology phenotypes. Currently,
no comprehensive ontology of physiology phenotypes is available. Our
work brings the immediate benefit that the availability of CPO enables
annotation of cellular phenotypes in the systems microscropy
community, and our method can be applied to physiology phenotypes in
other domains, in particular computational modelling.

Reviewer 2:
Minor revision (page 2 line 27): You introduce a Heart beating example
to illustrate the challenges in modeling processes and their
regulation. [...] The Heart example is illustrative but limited with
respect to your ontology. There, both the the function of the single
entity can be impaired and the regulation of multiple occurences of
processes (where the example ends due to the single heartedness of
many animals).

Response: 
In response to this comment as well as the similar comment reviewer 1 made, we have removed
the Heart beating example from the introduction and focus on the
detailed presentation of the processes involved in the Results section
of our manuscript.

@@

Reviewer 2:
Minor revision (page 3 second column line 33) see as above on the
phases of a single Heart action. The heart has no phase of
“inactivity” (that would be death), only a number of phases of which
the filling/ ejection phases are the least granular.

Response: We agree with the reviewer and we have changed our manuscript, according to his/her
suggestion.

@@

Reviewer 2:
Minor revision (page 5 Table 1): make it clear that you refer to the
product of row and column quantitites in the cells. I find it
difficult to understand this easy relation from your legend. Do you
really need this table?

Response:
We thank the reviewer for pointing this out and in response, we have removed the table from our manuscript. In the revised manuscript, we discuss the
content of the table in the main text.

### 2012-04-05 paper/drinking-bioinformatics.tex

Single occurrences of processes can be abnormal in multiple ways,
depending on the type of process.
%
First, common to all processes is the quality of {\em duration} and
consequently, each process can have an {\em abnormal} (increased or
decreased) duration. For example, a part of an organism may
participate in an {\em Inflammatory response} ({\tt GO:0006954}) that
lasts longer than normal, i.e., the organism has an {\em Increased
  duration of inflammatory response} phenotype. We define such a
phenotype as a phenotype of an organism which has a part that
participates in {\em Inflammatory response}, and this {\em
  Inflammatory response} process has an {\em Increased duration} ({\tt
  PATO:0000498}).

The second common type of abnormality are abnormalities based on
process participants in relation to the duration of the process. These
include all kinds of {\em rates} such as {\em mass flow rate}, {\em
  energy flow rate} and {\em velocity} (the rate of change of
position). In each of these cases, an object participates in a process
and a quality (or change of quality) of that object throughout the
duration of the process is considered to form a new quality. If the
process has participants that are distinguished into {\em inputs} and
{\em outputs}, then a recurring pattern is that the amount of inputs
or outputs that participate in the process can be {\em increased} or
{\em decreased}. For example, an {\em Increased rate of cytoplasmic
  streaming} can be defined as an increased amount of inputs or an
increased amount of outputs of a {\em cytoplasmic streaming} process.

@@

\bibliographystyle{plain}

\bibliography{/home/leechuck/Documents/papers/bibtex/lc}

>>>>

Single occurrences of processes can be abnormal in multiple ways,
depending on the type of process.
%
First, common to all processes is the quality of {\em Duration} and
consequently, each process can have an {\em abnormal} (increased or
decreased) duration. For example, a part of an organism may
participate in an {\em Inflammatory response} ({\tt GO:0006954}) that
lasts longer than normal, i.e., the organism has an {\em Increased
  duration of inflammatory response} phenotype. We define such a
phenotype as a phenotype of an organism which has a part that
participates in {\em Inflammatory response}, and this {\em
  Inflammatory response} process has an {\em Increased duration} ({\tt
  PATO:0000498}).

The second common type of abnormality are abnormalities based on
process participants in relation to the duration of the process. These
include all kinds of {\em rates} such as {\em Mass flow rate}, {\em
  Energy flow rate} and {\em Velocity} (the rate of change of
position). In each of these cases, an object participates in a process
and a quality (or change of quality) of that object throughout the
duration of the process is considered to form a new quality. If the
process has participants that are distinguished into {\em inputs} and
{\em outputs}, then a recurring pattern is that the amount of inputs
or outputs that participate in the process can be {\em increased} or
{\em decreased}. For example, an {\em Increased rate of cytoplasmic
  streaming} can be defined as an increased amount of inputs or an
increased amount of outputs of a {\em Cytoplasmic streaming} process.

@@

% \bibliographystyle{natbib}

% \bibliography{/home/leechuck/Documents/papers/bibtex/lc}
\begin{thebibliography}{}

\bibitem[Ashburner {\em et~al.}(2000)Ashburner {\em
  et~al.}]{Ashburner2000short}
Ashburner, M. {\em et~al.} (2000).
\newblock Gene ontology: tool for the unification of biology.
\newblock {\em Nature Genetics\/}, {\bf 25}(1).

\bibitem[Bada and Hunter(2007)Bada and Hunter]{Bada2007a}
Bada, M. and Hunter, L. (2007).
\newblock Enrichment of obo ontologies.
\newblock {\em Journal of Biomedical Informatics\/}, {\bf 40}(3), 300--315.

\bibitem[Bada {\em et~al.}(2004)Bada, Stevens, Goble, Gil, Ashburner, Blake,
  Cherry, Harris, and Lewis]{Bada2004}
Bada, M., Stevens, R., Goble, C., Gil, Y., Ashburner, M., Blake, J.~A., Cherry,
  M.~J., Harris, M., and Lewis, S. (2004).
\newblock A short study on the success of the gene ontology.
\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
  Web\/}, {\bf 1}(2), 235--240.

\bibitem[Bard {\em et~al.}(2005)Bard, Rhee, and Ashburner]{Bard2005}
Bard, J., Rhee, S.~Y., and Ashburner, M. (2005).
\newblock An ontology for cell types.
\newblock {\em Genome Biology\/}, {\bf 6}(2).

\bibitem[Berners-Lee {\em et~al.}(2001)Berners-Lee, Hendler, Lassila, {\em
  et~al.}]{Berners-Lee2001}
Berners-Lee, T., Hendler, J., Lassila, O., {\em et~al.} (2001).
\newblock {The Semantic Web}.
\newblock {\em Scientific American\/}, {\bf 284}(5), 28--37.

\bibitem[Conrad {\em et~al.}(2011)Conrad, W\"{u}nsche, Tan, Bulkescher,
  Sieckmann, Verissimo, Edelstein, Walter, Liebel, Pepperkok, and
  Ellenberg]{Conrad2011}
Conrad, C., W\"{u}nsche, A., Tan, T. H.~H., Bulkescher, J., Sieckmann, F.,
  Verissimo, F., Edelstein, A., Walter, T., Liebel, U., Pepperkok, R., and
  Ellenberg, J. (2011).
\newblock Micropilot: automation of fluorescence microscopy-based imaging for
  systems biology.
\newblock {\em Nature methods\/}, {\bf 8}(3), 246--249.

\bibitem[Cook {\em et~al.}(2011)Cook, Bookstein, and Gennari]{opb}
Cook, D.~L., Bookstein, F.~L., and Gennari, J.~H. (2011).
\newblock Physical properties of biological entities: An introduction to the
  ontology of physics for biology.
\newblock {\em PLoS ONE\/}, {\bf 6}(12), e28708.

\bibitem[Eilbeck {\em et~al.}(2005)Eilbeck, Lewis, Mungall, Yandell, Stein,
  Durbin, and Ashburner]{Eilbeck2005}
Eilbeck, K., Lewis, S.~E., Mungall, C.~J., Yandell, M., Stein, L., Durbin, R.,
  and Ashburner, M. (2005).
\newblock The sequence ontology: A tool for the unification of genome
  annotations.
\newblock {\em Genome Biology\/}, {\bf 6}(R55).

\bibitem[Engel {\em et~al.}(2010)Engel {\em et~al.}]{aposhort}
Engel, S.~R. {\em et~al.} (2010).
\newblock Saccharomyces genome database provides mutant phenotype data.
\newblock {\em Nucleic acids research\/}, {\bf 38}(Database issue).

\bibitem[Fuchs {\em et~al.}(2010)Fuchs, Pau, Kranz, Sklyar, Budjan, Steinbrink,
  Horn, Pedal, Huber, and Boutros]{Fuchs2010}
Fuchs, F., Pau, G., Kranz, D., Sklyar, O., Budjan, C., Steinbrink, S., Horn,
  T., Pedal, A., Huber, W., and Boutros, M. (2010).
\newblock Clustering phenotype populations by genome-wide {RNAi} and
  multiparametric imaging.
\newblock {\em Molecular Systems Biology\/}, {\bf 6}.

\bibitem[{Gene Ontology Consortium}(2010){Gene Ontology Consortium}]{go2010}
{Gene Ontology Consortium} (2010).
\newblock The gene ontology in 2010: extensions and refinements.
\newblock {\em Nucleic acids research\/}, {\bf 38}(Database issue), D331--335.

\bibitem[Gkoutos and Hoehndorf(2011)Gkoutos and Hoehndorf]{obml2011h1}
Gkoutos, G.~V. and Hoehndorf, R. (2011).
\newblock Ontology-based cross-species integration and analysis of
  saccharomyces cerevisiae phenotypes.
\newblock In {\em Proceedings of the 3rd Workshop for Ontologies in Biomedicine
  and Life sciences (OBML)\/}.

\bibitem[Gkoutos {\em et~al.}(2005)Gkoutos, Green, Mallon, Hancock, and
  Davidson]{Gkoutos2005}
Gkoutos, G.~V., Green, E.~C., Mallon, A.-M.~M., Hancock, J.~M., and Davidson,
  D. (2005).
\newblock {Using ontologies to describe mouse phenotypes.}
\newblock {\em Genome biology\/}, {\bf 6}(1).

\bibitem[Gkoutos {\em et~al.}(2009)Gkoutos, Mungall, Dolken, Ashburner, Lewis,
  Hancock, Schofield, Kohler, and Robinson]{Gkoutos2009b}
Gkoutos, G.~V., Mungall, C., Dolken, S., Ashburner, M., Lewis, S., Hancock, J.,
  Schofield, P., Kohler, S., and Robinson, P.~N. (2009).
\newblock Entity/quality-based logical definitions for the human skeletal
  phenome using {PATO}.
\newblock {\em Annual International Conference of the IEEE Engineering in
  Medicine and Biology Society.}, {\bf 1}, 7069--7072.

\bibitem[Goble and Stevens(2008)Goble and Stevens]{goble}
Goble, C. and Stevens, R. (2008).
\newblock State of the nation in data integration for bioinformatics.
\newblock {\em Journal of Biomedical Informatics\/}, {\bf 41}(5), 687--693.

\bibitem[Grau {\em et~al.}(2008)Grau, Horrocks, Motik, Parsia, Patelschneider,
  and Sattler]{Grau2008}
Grau, B., Horrocks, I., Motik, B., Parsia, B., Patelschneider, P., and Sattler,
  U. (2008).
\newblock {OWL} 2: The next step for {OWL}.
\newblock {\em Web Semantics: Science, Services and Agents on the World Wide
  Web\/}, {\bf 6}(4), 309--322.

\bibitem[Gruber(1995)Gruber]{Gruber1995}
Gruber, T.~R. (1995).
\newblock Toward principles for the design of ontologies used for knowledge
  sharing.
\newblock {\em International Journal of Human-Computer Studies\/}, {\bf
  43}(5-6).

\bibitem[Guarino(1998)Guarino]{Guarino1998}
Guarino, N. (1998).
\newblock Formal ontology and information systems.
\newblock In N.~Guarino, editor, {\em Proceedings of the 1st International
  Conference on Formal Ontologies in Information Systems\/}, pages 3--15. IOS
  Press.

\bibitem[Herre {\em et~al.}(2006)Herre, Heller, Burek, Hoehndorf, Loebe, and
  Michalek]{Herre2006}
Herre, H., Heller, B., Burek, P., Hoehndorf, R., Loebe, F., and Michalek, H.
  (2006).
\newblock {G}eneral {F}ormal {O}ntology ({GFO}) -- {A} foundational ontology
  integrating objects and processes [{V}ersion 1.0].
\newblock Onto-Med Report~8, IMISE, University of Leipzig, Leipzig, Germany.

\bibitem[Hoehndorf {\em et~al.}(2010a)Hoehndorf, Oellrich, and
  Rebholz-Schuhmann]{Hoehndorf2010phene}
Hoehndorf, R., Oellrich, A., and Rebholz-Schuhmann, D. (2010a).
\newblock Interoperability between phenotype and anatomy ontologies.
\newblock {\em Bioinformatics\/}, {\bf 26}(24), 3112 -- 3118.

\bibitem[Hoehndorf {\em et~al.}(2010b)Hoehndorf, Oellrich, Dumontier, Kelso,
  Rebholz-Schuhmann, and Herre]{Hoehndorf2010patterns}
Hoehndorf, R., Oellrich, A., Dumontier, M., Kelso, J., Rebholz-Schuhmann, D.,
  and Herre, H. (2010b).
\newblock Relations as patterns: Bridging the gap between {OBO} and {OWL}.
\newblock {\em BMC Bioinformatics\/}, {\bf 11}(1), 441+.

\bibitem[Horridge {\em et~al.}(2007)Horridge, Bechhofer, and
  Noppens]{Horridge2007}
Horridge, M., Bechhofer, S., and Noppens, O. (2007).
\newblock Igniting the {OWL} 1.1 touch paper: The {OWL} {API}.
\newblock In {\em Proceedings of OWLED 2007: Third International Workshop on
  OWL Experiences and Directions\/}.

\bibitem[Horrocks(2007)Horrocks]{Horrocks2007}
Horrocks, I. (2007).
\newblock {OBO} flat file format syntax and semantics and mapping to {OWL}
  {W}eb {O}ntology {L}anguage.
\newblock Technical report, University of Manchester.

\bibitem[Kazakov {\em et~al.}(2011)Kazakov, Kr{\"o}tzsch, and
  Siman\v{c}\'{i}k]{Kazakov2011}
Kazakov, Y., Kr{\"o}tzsch, M., and Siman\v{c}\'{i}k, F. (2011).
\newblock Unchain my $\mathcal{EL}$ reasoner.
\newblock In {\em Proceedings of the 23rd International Workshop on Description
  Logics (DL'10)\/}, CEUR Workshop Proceedings. CEUR-WS.org.

\bibitem[Lock and Strömblad(2010)Lock and Strömblad]{Lock2010}
Lock, J.~G. and Strömblad, S. (2010).
\newblock Systems microscopy: an emerging strategy for the life sciences.
\newblock {\em Experimental Cell Research\/}, {\bf 316}(8), 1438--1444.

\bibitem[Motik {\em et~al.}(2009)Motik, Grau, Horrocks, Wu, Fokoue, and
  Lutz]{owlprofiles}
Motik, B., Grau, B.~C., Horrocks, I., Wu, Z., Fokoue, A., and Lutz, C. (2009).
\newblock Owl 2 web ontology language: Profiles.
\newblock Recommendation, World Wide Web Consortium (W3C).

\bibitem[Mungall {\em et~al.}(2010a)Mungall, Gkoutos, Smith, Haendel, Lewis,
  and Ashburner]{Mungall2010}
Mungall, C., Gkoutos, G., Smith, C., Haendel, M., Lewis, S., and Ashburner, M.
  (2010a).
\newblock Integrating phenotype ontologies across multiple species.
\newblock {\em Genome Biology\/}, {\bf 11}(1), R2+.

\bibitem[Mungall {\em et~al.}(2010b)Mungall, Bada, Berardini, Deegan, Ireland,
  Harris, Hill, and Lomax]{Mungall2010go}
Mungall, C.~J., Bada, M., Berardini, T.~Z., Deegan, J., Ireland, A., Harris,
  M.~A., Hill, D.~P., and Lomax, J. (2010b).
\newblock Cross-product extensions of the gene ontology.
\newblock {\em Journal of biomedical informatics\/}.
\newblock in press.

\bibitem[Neumann {\em et~al.}(2010)Neumann, Walter, Hériché, Bulkescher,
  Erfle, Conrad, Rogers, Poser, Held, Liebel, and et~al.]{Neumann2010}
Neumann, B., Walter, T., Hériché, J.-K., Bulkescher, J., Erfle, H., Conrad,
  C., Rogers, P., Poser, I., Held, M., Liebel, U., and et~al. (2010).
\newblock Phenotypic profiling of the human genome by time-lapse microscopy
  reveals cell division genes.
\newblock {\em Nature\/}, {\bf 464}(7289), 721--727.

\bibitem[Ogren {\em et~al.}(2004)Ogren, Cohen, Acquaah-Mensah, Eberlein, and
  Hunter]{Ogren2004}
Ogren, P.~V., Cohen, K.~B., Acquaah-Mensah, G.~K., Eberlein, J., and Hunter, L.
  (2004).
\newblock The compositional structure of gene ontology terms.
\newblock {\em Pac Symp Biocomput\/}, pages 214--225.

\bibitem[\"{O}zg\"{o}vde and Gr\"{u}ninger(2010)\"{O}zg\"{o}vde and
  Gr\"{u}ninger]{Gruninger2010}
\"{O}zg\"{o}vde, A. and Gr\"{u}ninger, M. (2010).
\newblock Foundational process relations in bio-ontologies.
\newblock In {\em Proceeding of the 2010 conference on Formal Ontology in
  Information Systems\/}, pages 243--256, Amsterdam, The Netherlands, The
  Netherlands. IOS Press.

\bibitem[Robinson {\em et~al.}(2008)Robinson, Koehler, Bauer, Seelow, Horn, and
  Mundlos]{Robinson2008}
Robinson, P.~N., Koehler, S., Bauer, S., Seelow, D., Horn, D., and Mundlos, S.
  (2008).
\newblock The human phenotype ontology: a tool for annotating and analyzing
  human hereditary disease.
\newblock {\em American journal of human genetics\/}, {\bf 83}(5), 610--615.

\bibitem[Schindelman {\em et~al.}(2011)Schindelman, Fernandes, Bastiani, Yook,
  and Sternberg]{wpo}
Schindelman, G., Fernandes, J., Bastiani, C., Yook, K., and Sternberg, P.
  (2011).
\newblock Worm phenotype ontology: integrating phenotype data within and beyond
  the c. elegans community.
\newblock {\em BMC Bioinformatics\/}, {\bf 12}(1), 32.

\bibitem[Schmitz {\em et~al.}(2010)Schmitz {\em et~al.}]{Schmitz2010short}
Schmitz, M. H.~A. {\em et~al.} (2010).
\newblock Live-cell imaging rnai screen identifies pp2a–b55$\alpha$ and
  importin-$\beta$1 as key mitotic exit regulators in human cells.
\newblock {\em Nature Cell Biology\/}, {\bf 12}, 886--893.

\bibitem[Schofield {\em et~al.}(2011)Schofield, Sundberg, Hoehndorf, and
  Gkoutos]{Schofield2011}
Schofield, P.~N., Sundberg, J.~P., Hoehndorf, R., and Gkoutos, G.~V. (2011).
\newblock New approaches to the representation and analysis of phenotype
  knowledge in human diseases and their animal models.
\newblock {\em Briefings in Functional Genomics\/}, {\bf 10}(5), 258--265.

\bibitem[Smith {\em et~al.}(2007)Smith {\em et~al.}]{Smith2007short}
Smith, B. {\em et~al.} (2007).
\newblock The {OBO} {F}oundry: coordinated evolution of ontologies to support
  biomedical data integration.
\newblock {\em Nat Biotech\/}, {\bf 25}(11), 1251--1255.

\bibitem[Smith {\em et~al.}(2004)Smith, Goldsmith, and Eppig]{Smith2004}
Smith, C.~L., Goldsmith, C.-A.~W., and Eppig, J.~T. (2004).
\newblock The mammalian phenotype ontology as a tool for annotating, analyzing
  and comparing phenotypic information.
\newblock {\em Genome Biology\/}, {\bf 6}(1), R7.

\end{thebibliography}
